\chapter{Modeling}\label{chap:mod}

%In this chapter, you will learn how to create a problem

\section{Introduction}

The purpose of this chapter is to show how to create and manipulate
objects corresponding to the mathematical concepts of
{\it variable}, {\it function}, {\it constraint} and {\it system}.

We talk about {\it modeling} as these objects mathematically model the concrete problem you are faced to.

\subsection{C++ versus \quimper}

There are two possible alternatives for modeling.
You can either:
\begin{enumerate}
\item write C++ code. Variables, functions, constraints
and systems are C++ object that you declare yourself
and build by calling the constructors of the corresponding classes
\item write all these basic mathematic data in a text file, following
the (very intuitive) \quimper syntax. All these data are loaded
simultaneously and stored in a single \hcf{System} object.
\end{enumerate}

In both cases, you will access and use the data in the same way.
For instance, you will calculate the interval derivative of a function
by the same code, would it be created in your C++ program or loaded
from a \quimper file.

The chapter is organized as follows: we present
each concept (variable, function, etc.) in turn and
each time explain how objects are created in C++.

All the \quimper syntax is given afterwards, in a separate section.
 
\subsection{What we mean by ``variable'' and ``function''}

Let us rule out a potential ambiguity.

Since we are in the C++ programming language, the term {\it variable} and
{\it function} already refers to something precise. For instance, the following
piece of code introduces a {\it function} \cf{sum} and a {\it variable} \cf{x}:
\begin{lstlisting}
int sum(int x, int y) { 
  return x+y;
}

int x=2;
\end{lstlisting}
The variable $x$ may represent, say, the balance of a bank account.
The account number is what we call the {\it semantic} of $x$, that is, what $x$ is supposed
to represent in the user's mind. So, on one side,
we have {\it what we write}, that is, a program with variables and functions,
 and on the other side, {\it what we represent}, that is, concepts 
like a bank account.

With \ibex, we write programs to represent mathematical concepts
that are also called {\it variables} and {\it functions}.
The mapping $(x,y)\mapsto \sin(x+y)$ is an example of function that
we want to represent. It shall not be confused with the function \cf{sum}
above.

To avoid ambiguity, we shall talk about {\it mathematical}
variables (resp. functions) versus {\it program} variables (resp. functions).
We will also use italic symbol like $x$ to denote a mathematical variable
and postscript symbols like \cf{x} for program variables.
In most of our discussions, variables and functions will refer
to the mathematical objects so that the mathematical meaning will be the implicit one. 

\subsection{Arguments versus variables}

A (mathematical) variable does not necessarily represent a single real value.
It can also be a vector, a matrix or an array-of-matrices.
One can, e.g., build the following function 
$$\begin{array}{cccc}
f: & \Rset^2\times\Rset^3 & \to & \Rset\\
   &  (x,y) & \mapsto & x_1\times y_1+x_2\times y_2 - x_3
\end{array}.$$

In this case, $x$ and $y$ are vector variables with 2 and 3 components respectively.

We see, at this point, that the term {\it variable} becomes ambiguous.
For instance, if I say that the function $f$ takes $2$ variables, we don't really know if
it means that the function takes two arguments (that might be vectors or matrices) or if
the total input size is a vector of $\Rset^2$.

For this reason, from now on, we will call {\bf argument} the formal parameters
or input symbols the function has been defined with and {\bf variable} a component of the
latters. 

Hence, the function $f$ in the previous paragraphs has two arguments, $x$ and $y$
and 5 variables $x_1$, $x_2$, $y_1$, $y_2$ and $y_3$.

Note that, as a consequence, variables are always real-valued.

\section{Arguments}

Before telling you which class represents the arguments of a function, let us say first that
this class does not play a big role.
%The reason is that variables are only used to build functions.
Indeed, the only purpose of declaring an argument $x$ in \ibex is
for building a function right after, like $f:x\mapsto x+1$.
Functions play, in contrast, a big role.

In other words, $x$ is nothing but a syntaxic leaf in the expression 
of the function. In particular, an argument is not a slot for
representing domain. % (as it could be the case). %This is an important difference with other 
E.g, if you want to calculate the range of $f$ for $x\in [0,1]$,
you just call a (program) function \cf{eval} with a plain box in argument.
It's just as if $f$ was the function that takes one argument and
increment it, whatever the name of this argument is.

Once $f$ has been built, we can almost say that $x$ is no longer useful.
Arguments must be seen only as temporary objects, in the process of function construction.

Before going on, let us slightly moderate this point.
We have assumed here that, as a user of \ibex the operations you are interested in are: {\it evaluate $f$ on a box, 
calculate $f'$ on a box, solve $f(x)=0$} and so on. All these operations can be qualified as numerical: they take
intervals and return intervals. You don't need to deal again with the expression of the function, once built.
But if you need to handle, for any reason, the symbolic form of the function then you have to inspect the syntax
and arguments appear again.

\subsection{Dimensions and ordering}

We have said in the previous paragraph that an argument $x$ can actually represent
$n$ variables $x_1,\ldots,x_n$. So each argument has some associated information about
its dimension(s). 

Let us consider again this function: 
$$\begin{array}{cccc}
f: & \Rset^2\times\Rset^3 & \to & \Rset\\
   &  (x,y) & \mapsto & x_1\times y_1+x_2\times y_2 - x_3
\end{array}.$$

From the user standpoint, the function $f$ (once built) is ``flattened'' to a mapping from $\Rset^5$ to $\Rset$.
Each C++ function (eval, etc.) expects a 5-dimensional box as parameter.

The way intervals are mapped to the variables components follows a straightforward ordering:
everytime we call a (program) function of $f$ with the box $\ib=(\ib_1,\ldots,\ib_5)$ in argument, we simply enforce
$$x\in\ib_1\times\ib_2 \quad \mbox{and} \quad y\in\ib_3\times\ib_4\times\ib_5.$$

If you don't want to create functions in C++, you can skip now to Section \ref{sec:mod-func}.

\subsection{Class name and fields}

As we have just said, arguments are just symbols in expression. For this reason,
they are represented by a classe named \hcf{ExprSymbol}.
In fact, there is also another class we introduced for convenience, called \hcf{Variable}.
It is, of course, a very confusing name from the programer's viewpoint since a \cf{Variable}
does actually not represent a {\it variable} but an {\it argument}. However, from the user's viewpoint,
this distinction is not visible and ``variable'' is more meaningful than ``arguments''.
Anyway, the programer never has to deal with a \cf{Variable} object. Without going further into details, the \cf{Variable}
class must be seen as a kind of ``macro'' that generates \hcf{ExprSymbol} objects. 
This macro is only useful if you build arguments in C++ (see \S\ref{sec:mod-var-cpp}).

Once built, an argument is always typed \hcf{ExprSymbol}.

If \cf{x} is an \hcf{ExprSymbol} object, you can obtain the information about its dimensions via \cf{x.dim}.
The \hcf{dim} field is of type \hcf{Dim}, a class that simply contains 3 integers (one for each dimension, see
the API for further details).

Finally, an argument also has a name, that is only useful for displaying. It is a 
regular C string (\cf{char*}) stored in the field \hcf{name}.


%\subsection{Vector and matrix variables}



\subsection{Creating arguments (in C++)}\label{sec:mod-var-cpp}

The following piece of code creates an argument \cf{x} and prints it.

\begin{lstlisting}
  Variable x;
  cout << x << endl;
\end{lstlisting}

The first instruction creates a (program) variable \cf{x}. It is initialized by default, since
nothing is given here to the constructor.
By default, the argument is real (or {\it scalar}), meaning it is not a vector nor a matrix. 
Furthermore, the argument has a name that is automatically
generated. Of course, the name of the argument does not necessarily correspond to the name of the 
program variable.
For instance, \cf{x} is the name of a C++ variable but the corresponding argument is named {\it \_x\_0}.
The second instruction prints the name of the argument on the standard output:

\begin{lstlisting}
_x_0
\end{lstlisting}

It is possible to rename arguments, see below. %\S\ref{sec:mod-var-name}.

\subsubsection{Vector and matrix arguments}\label{sec:mod-var-vec}
To create a $n$-dimensional vector argument, just
give the number $n$ as a parameter to the constructor:

\begin{lstlisting}
  Variable y(3);   // creates a 3-dimensional vector
\end{lstlisting}

To create a $m\times n$ matrix, give $m$ (number of rows) and $n$ (number of columns) as parameters:

\begin{lstlisting}
  Variable z(2,3);   // creates a 2*3-dimensional matrix
\end{lstlisting}

We can go like this up to 3 dimensional arrays:

\begin{lstlisting}
  Variable t(2,3,4);   // creates a 2*3*4-dimensional array
\end{lstlisting}


\subsubsection{Renaming arguments}\label{sec:mod-var-name}
Usually, you don't really care about the names of arguments since you handle
program variables in your code.
However, if you want a more user-friendly display, you can specify
the name of the argument as a last parameter to the constructor.

In the following example, we create a scalar, a vector and a matrix argument each
time with a chosen name.

\begin{lstlisting}
  Variable x("x");     // creates a real argument named "x"
  Variable y(3,"y");   // creates a vector argument named "y"
  Variable z(2,3,"z"); // creates a matrix argument named "z"
  cout << x << " " << y << " " << z << endl;
\end{lstlisting}

Now, the display is:
\begin{lstlisting}
x y z
\end{lstlisting}


\section{Functions}\label{sec:mod-func}

Mathematical functions are represented by objects of the class \hcf{Function}.

These objects are very easy to build. Either refer to \S\ref{sec:mod-func-cpp}
or \S\ref{sec:mod-func-quimper}.

Various operations can be performed with a function. We detail below the main ones.

\subsection{Evaluation (forward computation)}

We show here how to calculate the image of a box by a function $f$ or, in short, how
to perform an {\it interval evaluation}.

We assume that the function \cf{f} has been created to represent $(x,y)\mapsto \sin(x+y)$.

We start by building a box (as explained in Chapter \ref{chap:arith}).
The box must have as many components as the function has variables, here, 2.

Then we simply call \hcf{f.eval(...)} to get the image of the box by \cf{f}:

\begin{lstlisting}
  double _box[][2]= {{1,2},{3,4}};
  IntervalVector box(2,_box);         // create the box ([1,2];[3,4])

  cout << "f(" << box << ")=" << f.eval(box) << endl; 
\end{lstlisting}

The display is:
\begin{verbatim}
f(([1,2] ; [3,4]))=[-1,-0.27941549819892486095]
\end{verbatim}

\subsection{Projection (backward computation)}
	
Projection means that we contract the input box $\ix$
with respect to $f(x)=y$, where $y$ is any constant (real value, interval,
inteval vector, etc.).

The underlying algorithm is the famous {\it forward-backward} (alias \textsc{HC4Revise})
algorithm. It is quick since it runs in linear time w.r.t. the size
of the constraint syntax and optimal when arguments have all one occurrence
in this syntax.

Consider again \cf{f} representing $(x,y)\mapsto\sin(x+y)$.
We can contract $\ix=[1,2]\times[3,4]$ w.r.t.
to $f(x)=-1$:

\begin{lstlisting}
  double _box[][2]= {{1,2},{3,4}};
  IntervalVector box(2,_box);
  cout << "initial box=" << box << endl;
  f.proj(-1.0,box);
  cout << "box after proj=" << box << endl;
\end{lstlisting}
The display is:
\begin{verbatim}
initial box=([1,2] ; [3,4])
box after proj=([1,1.7123889803846914504] ; [3,3.7123889803846914504])
\end{verbatim}
and one can indeed check that the resulting box is a consistent narrowing
of the initial one.

\subsection{Gradient}

Consider $f:(x,y)\mapsto x\times y$.
The first and most simple way of calculating the gradient is:
\begin{lstlisting}
  double init_xy[][2] = { {1,2}, {3,4} };
  IntervalVector box(2,init_xy);
  cout << "gradient=" << f.gradient(box) << endl;
\end{lstlisting}

Since $\fP{f}{x}=y$ and $\fP{f}{y}=x$ we get:
\begin{verbatim}
gradient=([3,4] ; [1,2])
\end{verbatim}

In this first variant, the returned vector is a new object created each time the function is called.
When we have to compute many times different values of the gradient for the same function, we can
also build a vector once for all and ask the \cf{gradient} to store the result in this slot:

\begin{lstlisting}
  IntervalVector g(4);
  f.gradient(box,g);
  cout << "gradient=" << g << endl;
\end{lstlisting}

\subsection{Jacobian and Hansen's matrix}

The interval Jacobian matrix of a function $f$ on a box $\ix$ is
$$J=\begin{pmatrix}
\fP{f_1}{x_1}(\ix) & \ldots & \fP{f_1}{x_n}(\ix)\\
\vdots \\
\fP{f_m}{x_1}(\ix) & \ldots & \fP{f_m}{x_n}(\ix)\\
\end{pmatrix}$$

The interval Jacobian matrix is obtained exactly as for the gradient.
Just write \\
\centerline{\cf{f.jacobian(box)}}\\ to get an \cf{IntervalMatrix}
containing an enclosure of the Jacobian matrix of $f$ on the box in argument.

There is also a variant where the matrix is passed as parameter
(as for the gradient) in order to avoid allocating memory for the calculated matrix:\\
\centerline{\cf{f.jacobian(box,J)}.}

You can also compute with \ibex the ``Hansen matrix''. This matrix
is another {\it slope} matrix, thiner than the interval Jacobian (but slower to be calculated).
It is, for example, used inside the interval Newton operator.
The Hansen matrix corresponds to the following matrix, where $(x_1,\ldots,x_n)$ denotes
the midvector of $\ix$.

$$\begin{pmatrix}
\fP{f_1}{x_1}(\ix_1,x_2,\ldots,x_n) & \fP{f_1}{x_2}(\ix_1,\ix_2,\ldots,x_n) & \ldots & \fP{f_1}{x_n}(\ix_1,\ix_2,\ldots,\ix_n) \\
\vdots & \\
\fP{f_m}{x_1}(\ix_1,x_2,\ldots,x_n) & \fP{f_n}{x_2}(\ix_1,\ix_2,\ldots,x_n) & \ldots & \fP{f_m}{x_n}(\ix_1,\ix_2,\ldots,\ix_n) \\
\end{pmatrix}$$

\subsection{Creating functions (in C++)}\label{sec:mod-func-cpp}

The following piece of code creates the function
$(x,y)\mapsto \sin(x+y)$:

\begin{lstlisting}	
  Variable x("x");
  Variable y("y");
  Function f(x,y,sin(x+y));
  cout << f << endl;
\end{lstlisting}

The display is:
\begin{lstlisting}
_f_0:(x,y)->sin((x+y))
\end{lstlisting}
%_f_0:(_x_0,_x_1)->sin((_x_0+_x_1))

You can directly give up to 6 variables in argument of the \cf{Function} constructor:
\begin{lstlisting}
Variable a,b,c,d,e,f,g;
Function f(a,b,c,d,e,f,g,a+b+c+d+e+f+g);
\end{lstlisting}	

If more than 6 variables are needed, you need to build an intermediate array for collecting the arguments.
More precisely, this intermediate object is an \cf{Array<const ExprSymbol>}. The usage is summarized below. In this
example, we have 7 variables. But instead of creating the function
$$x\mapsto x_1+\ldots+x_7$$
with one argument (a vector with 7 components), we decide to create the function
$$(x_1,\ldots,x_7)\mapsto x_1+\ldots+x_7.$$
with 7 arguments (7 scalar variables).
\begin{lstlisting}
Variable x[7]; // not to be confused with x(7)
Array<const ExprSymbol> vars(7);
for (int i=0; i<7; i++)
  vars.set_ref(i,x[i]);
Function f(vars, x[0]+x[1]+x[2]+x[3]+x[4]+x[5]+x[6]);
cout << f << endl;
\end{lstlisting}	

The display is:
\begin{verbatim}
_f_1:(_x_0,_x_1,_x_2,_x_3,_x_4,_x_5,_x_6)->
     ((((((_x_0+_x_1)+_x_2)+_x_3)+_x_4)+_x_5)+_x_6)
\end{verbatim}

\subsubsection{Renaming functions}

By default, function names are also generated. But you can also set your own function name, as the last parameter of the constructor:
\begin{lstlisting}
Function f(x,y,sin(x+y),"f");
\end{lstlisting}

\subsubsection{Allowed symbols}

The following symbols are allowed in expressions:
\begin{verbatim}
sign, min, max,
sqr, sqrt, exp, log, pow, 
cos, sin, tan, acos, asin, atan,
cosh, sinh, tanh, acosh, asinh, atanh
atan2
\end{verbatim}

Power symbols \cf{^} are not allowed. You must
either use \cf{pow(x,y)}, or simply \cf{sqr(x)} for the square function.

Here is an example of the distance function between (\cf{xa},\cf{ya}) and
(\cf{xb},\cf{yb}):

\begin{lstlisting}
  Variable xa,xb,ya,yb;
  Function dist(xa,xb,ya,yb, sqrt(sqr(xa-xb)+sqr(ya-yb)));
\end{lstlisting}

\subsubsection{Functions with vector arguments}

If arguments are vectors, you can refer to the component
of an argument using square brackets. Indices start by 0,
following the convention of the C language.

We rewrite here the previous distance function using 2-dimensional
arguments \cf{a} and \cf{b} instead:
\begin{lstlisting}
  Variable a(2);
  Variable b(2);
  Function dist(a,b,sqrt(sqr(a[0]-b[0])+sqr(a[1]-b[1])),"dist");
\end{lstlisting}

\subsubsection{Composition}

You can compose functions. Each argument of the called function can be substitued
by an argument of the calling function, a subexpression or a constant value.

For instance, we can define a function "foo" that associates to
a point \cf{x} the distance between \cf{x} and a fixed point $(1,2)$,
using the generic distance function defined in the previous paragraph:

\begin{lstlisting}
  Vector pt(2);
  pt[0]=1;
  pt[1]=2;

  Variable x(2);
  Function f(x,dist(x,pt),"foo");
\end{lstlisting}

\subsubsection{Vector-valued functions}

To define a vector-valued function, the \hcf{Return} keword allows
you to list the function's components.

For instance, we can define the function that associates to $x$ the 
respective distances between two fixed points \cf{pt1}=$(0,0)$ and \cf{pt2}=$(1,1)$.

\begin{lstlisting}	
	Variable x(2,"x");
	Variable pt(2,"p");
	Function dist(x,pt,sqrt(sqr(x[0]-pt[0])+sqr(x[1]-pt[1])),"dist");

	Vector pt1=Vector::zeros(2);
	Vector pt2=Vector::ones(2);

	Function f(x,Return(dist(x,pt1),dist(x,pt2)),"f");

	cout << f << endl;
\end{lstlisting}

The display is as folllows. Note that constant values like 0 are automatically replaced
by degenerated intervals (like [0,0]):
\begin{verbatim}
f:(x)->(dist(x,([0,0] ; [0,0])),dist(x,([1,1] ; [1,1])))
\end{verbatim}

\section{Constraints}

In this section, we do not present {\it constraints} in their full generality
but {\it numerical constraints} (the ones you are the most likely interested in).

A numerical constraint in \ibex is either a relation like 
$f(x)<0$, $f(x)\le0$, $f(x)=0$, $f(x)\ge0$ or $f(x)>0$,
where $f$ is a function as introduced in the previous section. If $f$ is vector-valued, then
$0$ must be a vector.

Surprisingly, constraints do not play an important role in \ibex.
It sounds a little bit contraditory for a {\it constraint} programming library.
The point is that \ibex is rather a {\it contractor} programming library meaning that 
we build, apply and compose contractors rather than constraints directly.

As a programer, you may actually face two different situations.

Either you indeed want to use a constraint as a contractor in which case you
build a \cf{Contractor} object with this constraint (the actual class depending
on the algorithm you chose, as detailed in Chapter \ref{chap:ctc} --by default, it is \cf{HC4Revise}--).
Either you need to do something else, say, like calculating the Jacobian matrix of the
function $f$. In this case, you just need to get a reference to this function 
and call \cf{jacobian}. In fact, all the information inherent to a constraint
(except the comparison operator of course) is contained in the underlying function so that there
is little specific code related to the constraint itself.

For these reasons, the only operations you actually do with a constraint is either
to read its field or wrap it into a contractor.

\subsection{Class and Fields}

The class for representing a numerical constraint is \hcf{NumConstraint}.
The first field in this class is a reference to the function:\\
\centerline{\cf{Function&} \hcf{f}}

The second field is the comparison operator:\\
\centerline{\cf{CmpOp} \hcf{op}}

\cf{CmpOp} is just an \cf{enum} (integer) with the following values

\begin{center}
\begin{tabular}{cccc}
\hcf{LT} & $<$ & \hcf{LEQ} & $\le$ \\
\hcf{EQ} & =& \\
\hcf{GEQ} & $\ge$ & \hcf{GT} & $>$ \\
\end{tabular}
\end{center}

\subsection{Creating constraints (in C++)}\label{sec:mod-ctr-cpp}

To create a numerical constraint, you can build the function $f$ first and
call the constructor of \cf{NumConstraint} as in the following example.

\begin{lstlisting}
Variable x;
Function f(x,x+1);
NumConstraint c(f,LEQ); // the constraint x+1<=0
\end{lstlisting}	

But you can also write directly:
\begin{lstlisting}
Variable x;
NumConstraint c(x,x+1<=0);
\end{lstlisting}	
which gives the same result. The only difference is that, in the second case,
the object \cf{c.f} is ``owned'' (and destroyed) by the constraint whereas 
in the first case, \cf{c.f} is only a reference to \cf{f}.

Note that the constant $0$ is automatically interpreted as a vector (resp. matrix),
if the left-hand side expression is a vector (resp. matrix). However, it does not
work for other constants: you have to build the constant with the proper dimension,
e.g.,
\begin{lstlisting}
Variable x(2);
NumConstraint c(x,x=IntervalVector(2,1)); // the constraint x=(1,1)
cout << "c=" << c << endl;
\end{lstlisting}	

The display is:
\begin{verbatim}
c=(_x_0-([1,1] ; [1,1]))=0
\end{verbatim}

In case of several variables, the constructor of \cf{NumConstraint} works as for functions.
Up to 6 variables can be passed as arguments:
\begin{lstlisting}
Variable a,b,c,d,e,f,g;
NumConstraint c(a,b,c,d,e,f,g,a+b+c+d+e+f+g<=1);
\end{lstlisting}	

And if more variables are necessary, you need to build an \cf{Array<const ExprSymbol>} first.
See \S\ref{sec:mod-func-cpp}.

\section{Systems}

A {\it system} in \ibex is a set of constraints with, optionnaly, a goal function to minimize.
One is usually interested in solving the system while minimizing the criterion, if any.

For this reason, a system is not as simple as a collection of {\it any} constraints:
each constraint must exactly relates the same set of arguments. And this set must
also coincide with that of the goal function.
Many algorithms of \ibex are based on this assumption.
This is why they requires a system as argument (and not just an array of constraints).
This makes systems a central concept in \ibex.

A system is an object of the \hcf{System} class. This object is made of several fields
that are detailed below. %We take as example the following system:

\begin{itemize}
\item \cf{const int} \hcf{nb_var}: the total number of variables or, in other words, the
{\it size} of the problem. This number is basically the sum of all arguments' components. For instance,
if one declares an argument $x$ with 10 components and an argument $y$ with 5, the value of this field
will be $15$.
\item \cf{const int} \hcf{nb_ctr}: the number of constraints
\item \cf{Function*} \hcf{goal}: a pointer to the goal function. If there is no goal function, this
pointer is \cf{NULL}.
\item \cf{Function} \hcf{f}: the (usually vector-valued) function representing the constraints. 
For instance, if one defines three constraints: $x+y\leq0$, $x-y=1$ and $x-y\geq0$, the function f will be 
$(x,y)\mapsto (x+y,x-y-1,x-y)$. Note that the constraints are automatically transformed so that the right side 
is 0 but, however, without changing the comparison sign. It is however possible to {\it normalize} a system so that
all inequalities are defined with the $\le$ sign (see \S\ref{sec:mod-sys-transfo}).
\item \cf{IntervalVector} \hcf{box}: when a system is loaded from a file (see \S\ref{sec:mod-sys-load}),
a initial box can be specified. It is contained in this field.
\item \cf{Array<NumConstraint>} \hcf{ctrs}: the array of constraints. The \cf{Array} class of \ibex can
be used as a regular C array.
\end{itemize}

\subsection{Copy and transformation}\label{sec:mod-sys-transfo}

\subsection{Auxiliary functions}\label{sec:mod-sys-auxfunc}

\subsection{Creating systems (in C++)}

The first alternative for creating a system is to do it programmatically, that is, directly in your C++ program.
Creating a system in C++ resorts to a temporary object called a {\it system factory}. The task is done in a few simple steps:
\begin{enumerate}
\item declare a new system factory (an object of \hcf{SystemFactory})
\item add arguments in the factory using \hcf{add_var}.
\item (optional) add the expression of the goal function using \hcf{add_goal}
\item add the constraints using \hcf{add_ctr}
\item create the system simply by passing the factory to the constructor of \cf{System}
\end{enumerate}

Here is an example:

\begin{lstlisting}
  Variable x,y;

  SystemFactory fac;
  fac.add_var(x);
  fac.add_var(y);
  fac.add_goal(x+y);
  fac.add_ctr(sqr(x)+sqr(y)<=1);

  System sys(fac);
\end{lstlisting}

If you compare the declaration of the constraint here with the examples given in \S\ref{sec:mod-ctr-cpp},
you notice that we do not list here the arguments before writing {\tt sqr(x)+sqr(y)<=1}.
The reason is simply that, as said above, the goal function and the constraints in a system
share all the same list of arguments. This list is defined via \cf{add_var} once for all.

\section{The \quimper syntax}\label{sec:mod-sys-load}

Entering a system programmatically is usually not very convenient.
You may prefer to separate the model of the problem from the algorithms
you use to solve it. In this way, you can run the same program with different
variants of your model without recompiling it each time.

\ibex provides such possibility. You can directly load a system from a (plain text) input file.

Here is a simple example. Copy-paste the text above in a file named, say, {\tt problem.txt}. 
The syntax talks for itself:

\begin{verbatim}
Variables
  x,y;

Minimize
  x+y;

Constraints
  x^2+y^2<=1;
end
\end{verbatim}

Then, in your C++ program, just write:

\begin{lstlisting}
System sys("problem.txt");
\end{lstlisting}

and the system you get is exactly the same as in the previous example.

Next sections details the mini-language of these input files. 

\subsection{Overall structure}
First of all, the input file is a sequence of declaration blocks that must respect the following order:
\begin{enumerate}
\item (optional) constants
\item variables
\item (optional) auxiliary functions
\item (optional) goal function
\item constraints
\end{enumerate}

Next paragraph gives the basic format of numbers and intervals.
The subsequent paragraphs detail each declaration blocks.

\subsection{Real and Intervals}
A real is represented with the usual English format, that is
with a dot separating the integral from the decimal part,
and, possibly, using scientific notation.

Here are some valid examples of reals in the syntax:
\begin{verbatim}
0
3.14159
-0.0001
1.001e-10
+70.0000
\end{verbatim}
 
An interval are two reals separated by a comma
and surrounded by square brackets. The special symbol
{\tt oo} (two consecutive ``o'') represents the infinity $\infty$.
Note that, even with infinity bounds, the brackets
must be squared (and not parenthesis as it should be since the
bound is open). Here are some examples:

\begin{verbatim}
[0,1]
[0,+oo]
[-oo,oo]
[1.01e-02,1.02e-02]
\end{verbatim}

\subsection{Constants}
Constants are all defined in the same declaration block, 
started with the \cf{Constants} keyword.
A constant value can depends on other (previously defined) constants value. Example:

\begin{verbatim}
Constants
  pi=3.14159;
  y=-1.0;
  z=sin(pi*y);
\end{verbatim} 

You can give a constant an interval enclosure rather than a single fixed value.
This interval will be embedded in all subsequent computations.
Following the previous example, we can give \cf{pi} a valid enclosure as below.
We just have to replace ``{\tt =}'' by ``{\tt in}''.


\begin{verbatim}
Constants
  pi in [3.14159,3.14160];
  y=-1.0;
  z=sin(pi*y);
\end{verbatim}

Constants can also be vectors, matrices or array of matrices.
You need to specify the dimensions of the constant in square brackets.
For instance \cf{x} below is a column vector with 2 components, the first component is equal
to $0$ and the second to $1$:
\begin{verbatim}
Constants
x[2] = (0; 1);
\end{verbatim}
Writing {\tt x[2]} is equivalent to {\tt x[2][1]} because a column vector is also
a $2\times 1$ matrix.
A row vector is a $1\times 2$ matrix so a row vector has to be declared as follows. 
On the right side, note that we use commas instead of periods:
\begin{verbatim}
Constants
x[1][2] = (0, 1);
\end{verbatim}

{\bf important remark}. 
The reason why the syntax for declaring row vectors differs here from Matlab is that a 2-sized row vector surrounded
by brackets would conflict with an interval. So, do note confuse
\cf{[0,1]} with \cf{(0,1)}:
\begin{itemize}
\item \cf{(0,1)} is a 2-dimensional row vector of two reals, namely 0 and 1.
This is {\bf not} an open interval.
\item \cf{[0,1]} is the 1-dimensional interval $[0,1]$. This is {\bf not} a 2-dimensional row vector.
\end{itemize}
%The difference is ``\cf{;}'' versus ``\cf{,}''.

Of course, you mix vector with intervals. For instance:
{\tt ([-oo,0];[0,+oo])} is a column vector of $2$ intervals, $(-\infty,0]$ and $[0,+\infty)$.

Here is an example of matrix constant declaration: %, add an extra dimension in brackets:
\begin{verbatim}
Constants
M[3][2] = ((0 , 0) ; (0 , 1) ; (1 , 0));
\end{verbatim}
This will create the constant matrix \cf{M} with 3 rows and 2 columns equal to
$$\begin{pmatrix}
0 & 0 \\ 0 & 1 \\ 1 & 0
\end{pmatrix}$$

You can also declare array of matrices:
\begin{verbatim}
Constants
 c[2][2][3]=(((0,1,2); (3,4,5)) ; ((6,7,8); (9,10,11)));
\end{verbatim}
It is possible to define up to three dimensional vectors, but not more.

When all the components of a multi-dimensional constant share the same interval, you
don't need to duplicate it on the right side. Here is an example of a $10\times10$ matrix where all
components are $[0,0]$:
\begin{verbatim}
Constants
 c[10][10] in [0,0];
\end{verbatim}
Ibex intializes the 100 entries of the matrix {\tt c} to $[0,0]$.

Finally, the following table summarizes the possibility for declaring constants
through different examples.

\begin{tabular}{p{.3\textwidth}|p{.7\textwidth}}
\hline
{\tt x in [-oo,0]} & declares a constant {\tt x} $\in(-\infty,0]$ \\
{\tt x in [0,1]} & declares an constant {\tt x} $\in[0,1]$ \\
{\tt x in [0,0]} & declares a constant {\tt x} $\in[0,0]$ \\
{\tt x = 0} & declares a real constant {\tt x} equal to $0$ \\
{\tt x = 100*sin(0.1)} & declares a constant {\tt x} equal to $100\sin(0.1)$ \\
{\tt x[10] in [-oo,0]} & declares a $10$-sized constant vector {\tt x}, with each component {\tt x(i)} $\in(-\infty,0]$ \\
{\tt x[2] in ([-oo,0];[0,+oo])} & declares a $2$-sized constant vector {\tt x} with {\tt x(1)}$\in(-\infty,0]$ and {\tt x(2)}$\in[0,+\infty)$ \\
%{\tt x[10] in [0,1]} & declares $10$ constants $x[1],\ldots,x[10]$, each with domain $[0,1]$ \\
{\tt x[3][3] in} &  \multirow{4}{*}{declares a constrant matrix 
{\tt x} $\in\begin{pmatrix}
\!\,[0,1] & 0 & 0 \\
0 & [0,1]& 0 \\
0 & 0 & [0,1] 
\end{pmatrix}$.} \\
\!\,{\tt (([0,1],0,0);} & \\
\!\,{\tt (0,[0,1],0);} & \\
\!\,{\tt (0,0,[0,1]))} &\\
{\tt x[10][5] in [0,1]} & declares a matrix {\tt x} with each entry {\tt x(i,j)} $\in[0,1]$.\\
{\tt x[2][10][5] in [0,1]} & declares an array of 2 $10\times 5$ matrices with each entry {\tt x(i,j,k)} $\in[0,1]$.\\
\end{tabular}

\subsection{Functions}\label{sec:mod-func-quimper}

When the constraints involve the same expression repeatidly, it may be
convenient for you to put this expression once for all in a separate auxiliary
function and to call this function.

Assume for instance that your constraints intensively use the following expression
$$\sqrt{(x_a-x_b)^2+(y_a-y_b)^2)}$$
where $x_a,\ldots y_b$ are various sub-expressions, like:
\begin{verbatim}
sqrt((xA-1.0)^2+(yA-1.0)^2<=0;
sqrt((xA-(xB+xC))^2+(yA-(yB+yC))^2=0;
...
\end{verbatim}

You can declare the distance function as follows.
\begin{verbatim}
function distance(xa,ya,xb,yb)
 return sqrt((xa-xb)^2+(ya-yb)^2;
end
\end{verbatim}
%Note that \cf{d} corresponds to the returned value of the function,
%as in the Matlab syntax. We will call it the {\it return variable}.
You will then be able to simplify the writing of constraints:
\begin{verbatim}
distance(xA,1.0,yA,1.0)<=0;
distance(xA,xB+xC,yA,yB+yC)=0;
...
\end{verbatim}

As you may expect, this will result in the creation of
a \cf{Function} object (see \S\ref{sec:mod-func}) that
you can access from your C++ program. This will be explained
in \ref{sec:mod-sys-auxfunc}.

A function can return a single value, a vector
or a matrix. Similarly, it can take real, vectors or matrix arguments.
You can also write some minimal ``code'' inside the function before
returning the final expression.
%assigning the final expression to the return variable. 
This code is however limited to be a sequence of assignments.

Let us now illustrate all this with a more sophisticated example.
We write below the function that calculates the rotation matrix
from the three Euler angles, $\phi$, $\theta$ and $\psi$ :

$R : (\phi,\psi,\theta) \mapsto$
{\scriptsize
$$\begin{pmatrix}
\cos(\theta)\cos(\psi) & -\cos(\phi)\sin(\psi)+\sin(\theta)\cos(\psi)\sin(\phi) & \sin(\psi)\sin(\phi)+\sin(\theta)\cos(\psi)\cos(\phi)\\
\cos(\theta)\sin(\psi) & \cos(\psi)\cos(\phi)+\sin(\theta)\sin(\psi)\sin(\phi) & -\cos(\psi)\sin(\phi)+\sin(\theta)\cos(\phi)\sin(\psi)\\
-\sin(\theta) & \cos(\theta)\sin(\phi) & \cos(\theta)\cos(\phi);
\end{pmatrix}
$$}

As you can see, there are many occurrences of the same subexpression
like $\cos(\theta)$ so a good idea for both readibility and (actually) efficiency
is to precalculate such pattern and put the result into an intermediate variable.

Here is the way we propose to define this function:

\begin{verbatim}
/* Computes the rotation matrix from the Euler angles: 
   roll(phi), the pitch (theta) and the yaw (psi)  */
function euler(phi,theta,psi)
  cphi   = cos(phi);
  sphi   = sin(phi);
  ctheta = cos(theta);
  stheta = sin(theta);
  cpsi   = cos(psi);
  spsi   = sin(psi);
  
  return 
  ( (ctheta*cpsi, -cphi*spsi+stheta*cpsi*sphi, 
                   spsi*sphi+stheta*cpsi*cphi) ; 
    (ctheta*spsi, cpsi*cphi+stheta*spsi*sphi, 
                 -cpsi*sphi+stheta*cphi*spsi) ;
    (-stheta, ctheta*sphi, ctheta*cphi) );
end
\end{verbatim}

{\bf Remark}. Introducing temporary variables like \cf{cphi} amouts to build a DAG instead of
a tree for the function expression. It is also possible (and easy) to build a DAG when you directly create
a \cf{Function} object in C++ (to be documented).

\subsection{Variables}

Variables are defined exactly in the same fashion as constants (see \S\ref{ref:cst}).
It is possible to define up to three dimensional vectors, with an optional domain to initialize each
component with. The following examples are valid:

{\tt x[10][5][4];}

{\tt x[10][5][4] in [0,1];}

Whenever domains are not specified, they are set by default to $(-\infty,+\infty)$.

\subsection{Constraints}
\subsubsection{Loops}

You can resort to loops in a Matlab-like syntax to define constraints. Example:

\begin{verbatim}
Variables
  x[10];

Constraints
  for i=1:10;
    x[i] <= i;
  end
end
\end{verbatim}

\subsection{Difference with C++}

\begin{itemize}
\item Vectors are surrounded by parenthesis (not brackets)
\item Indices start by $1$ instead of $0$
\item You can use the \^{} symbol
\end{itemize}

